---
title: "Functional analyses"
subtitle: "Workshop on RNA-Seq"
author: "`r paste0('Paulo Czarnewski | <b>NBIS</b> | ',format(Sys.time(), '%d-%b-%Y'))`"
output:
  bookdown::html_document2:
          toc: true
          toc_float: true
          toc_depth: 4
          number_sections: true
          theme: flatly
          highlight: tango
          df_print: paged
          code_folding: "none"
          self_contained: false
          keep_md: false
          encoding: 'UTF-8'
          css: "assets/lab.css"
editor_options: 
  chunk_output_type: console
---

```{r,child="assets/header-lab.Rmd"}
```

<!-- ------------ Only edit title, subtitle & author above this ------------ -->

Loading packages and data

```{r}
#load the packages you need
library(pheatmap)
library(rafalib)
library(DESeq2)
library(pvclust)
library(biomaRt)
```

# Functional Analysis
## Overrepresentation analysis
### enrichR

```{r,results='hide',block.title=TRUE,fig.height=5,fig.width=10}

```

## GSA
### piano

```{r,results='hide',block.title=TRUE,fig.height=5,fig.width=10}

```

### Links to packages:

topGO  
goana  
topKEGG  
kegga  
romer  
piano  
roast  
fry  
camera  

### Links to webtools:

Package | link | Database | Comment  
GOrilla | http://cbl-gorilla.cs.technion.ac.il | GO | Support for REVIGO  
REVIGO | http://revigo.irb.hr | GO | Summarises redundancy  
Enrichr | http://amp.pharm.mssm.edu/Enrichr/ | GO, KEGG, TF,  many others  | Extensive   
DAVID | https://david.ncifcrf.gov | GO, KEGG, TF, many others | *Not updated  
KEGG | https://www.genome.jp/kegg/ | KEGG | Shows the pathways  

***
## GSEA

```{r,results='hide',block.title=TRUE,fig.height=5,fig.width=10}

```


***
## Cell-type enrichment / deconvolution

* xCell

```{r,results='hide',block.title=TRUE,fig.height=5,fig.width=10}

```

***
## Protein-protein interactions

### Links to webtools:

* GeneMANIA: https://genemania.org
* STRING-DB: https://string-db.org
* MIST: https://fgrtools.hms.harvard.edu/MIST/help.jsp

__End of document__
